---
title: "fuzzy-clustering"
output:
  rmarkdown::html_vignette:
    number_sections: true
vignette: >
  %\VignetteIndexEntry{fuzzy-clustering}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(fussclust)
```


# Description

The `fussclust` package provides methods for distance-based fuzzy clustering:
both unsupervised and semi-supervised.
The core need for the package was to provide semi-supervised models in \texttt{R},
as there lacked a proper implementation of these, but in fact - 
both unsupervised and semi-supervised fuzzy clustering follow the same problem
formulation and solution.
Hence, `fussclust` implements a common estimation platform for the unified class
of *fuzzy clustering* (unsupervised and semi-supervised).

With the current release, `fussclust` enables four classical models:

* (unsupervised) Fuzzy c-Means,
* Semi-Supervised Fuzzy c-Means,
* Possibilistic c-Means,
* Semi-Supervised Possibilistic c-Means.

The typical workflow is:

1. prepare your *features matrix* `X` in a form of a numerical matrix,
2. if running semi-supervised models, prepare partial supervision in a form
of a binary matrix `superF` (see details in Section 4.1),
3. fit the appropriate selected model,
4. check the outputs: estimated memberships `U`, or estimated prototypes `V`.

# Methodology

Fuzzy clustering covered in `fussclust` is distance-based.
The soft memberships of $N$ objects to $c$ clusters are estimated based on
an assumption that similarity can be approximated by distances - objects that
are close in a given metric space are considered similar.

The optimization problem is formulated as

$$\min \: Q(U,V; X)$$

where $Q$ is an objective function, $U$ is the memberships matrix, 
$V$ are the cluster prototypes, and $X$ is a features matrix.

The problem does not have a closed solution, and therefore, the approximate
solution is used - the so-called Alternating Optimization algorith.
It relies on estimating separately optimal memberships $\hat{U}$ and optimal prototypes
$\hat{V}$ by considering partial functions $Q_U(V)$ and $Q_V(U)$, correspondingly.
For details, see [Kmita et al. (2024)](https://doi.org/10.1109/TFUZZ.2024.3370768).

The AO algorithm can be summarized as:

1. initiate algorithm with a proper initialization memberships $\hat{U}^{(0)}.
2. set loop counter $l = 1$.
3. estimate prototypes $\tilde{V}^{(l)} using the function $\hat{V}$,
4. estimate memberships $\tilde{U}^{(l)} using the function $\hat{U}$,
5. check convergence - if difference between $\tilde{U}^{(l)}$
and $\tilde{U}^{(l-1)}$ is less than a threshold, stop the algorithm.
Otherwise, go back to point 2, and increase the loop counter by $1$.

# Unsupervised models

We will provide walt-through examples based on the well-known `iris` dataset.
We will use two of the available features -- sepal length and sepal width --
in our examples.

## Fitting the model

Let us prepare the matrix.

```{r}
X <- iris[, c("Sepal.Length", "Sepal.Width")] |> as.matrix()

cat(paste0("Class of `iris` is ", class(iris), " and class of `X` is ", paste0(class(X), collapse = " & ")))
```

In case of `iris` dataset, we know there are three classes -- each class being
a different species of irises --, so we will decide to fit `C = 3` clusters.
Nevertheless, the user can choose any $C \leq 2$.

We fit Fuzzy c-Means model (`fussclust::FCM`) and Possibilistic c-Means model
(`fussclust::PCM`).

```{r}
model_fcm <- fussclust::FCM(X = X, C = 2)
model_pcm <- fussclust::PCM(X = X, C = 2)
```

and investigate the estimated prototypes from `model_fcm`, and estimated
memberships from `model_pcm` for ten objects.

```{r}
model_fcm$V
```

```{r}
model_pcm$U[1:10,]
```

# Semi-Supervised models

## Encoding partial supervision

As stated in Section 1 ''Description'', `fussclust` requires to encode partial
supervision in form of a binary matrix `superF`.

The assumption undertaken in `fussclust` is that the number of clusters $C$
is set to be the number of unique classes.

Typically, the class belonging information (partial supervision) is stored
in a vector called the response variable.
This is the case with iris as well:

```{r}
iris[1:3, ]
```
The information is stored in the `Species` column, with all unique species being:

```{r}
unique(iris$Species)
```

To encode such supervision, appropriate data processing is needed.
First, an order of the classes must be arbitrarily set by the user of the model.
Let $\mathcal{Y}$ be a $3$-tuple, introducing the order of species:

$$\mathcal{Y} = (y^1 = \texttt{setosa}, y^2 = \texttt{versicolor}, y^3 = \texttt{virginica})$$.

The `superF` is constructed in such a way.
Let $F = [f_{jk}]$ be the binary supervision matrix (mathematical symbol corresponding to `superF`),
where $j=1,\ldost,N$ indexes observations, and $k=1,\ldots,c$ indexes clusters.

* it has `nrow = nrow(X)`,
* it has $C$ columns,
* each $k$th column encodes information abouth $k$th class
(the corresponding $k$th element $y^k$ of $\mathcal{Y}$),
* $f_{jk} = 1$ only when the $j$th observation is assumed to belong to the $k$th class.

In case of the considered example, this can be implemented as follows.

```{r}
superF <- matrix(0, nrow = nrow(iris), ncol = 3)
superF[iris$Species == "setosa", 1] <- 1
superF[iris$Species == "versicolor", 2] <- 1
superF[iris$Species == "virginica", 3] <- 1
```

## Fitting the model

When fitting the model, the user must obligatory provide the value of 
the scaling factor $\alpha$ (`alpha`) regulating the impact of partial supervision.

This is a delicate matter and there can be no unequivocal guidance on a fit-for-all
value. For more details, see the discussion in 
[Kmita et al. (2024)](https://doi.org/10.1109/TFUZZ.2024.3370768).

For our example, we will choose $\alpha = 1$.

Semi-Supervised Fuzzy c-Means model:

```{r}
model_ssfcm <- fussclust::SSFCM(X = X, C = 3, alpha = 1, superF = superF)
```

Semi-Supervised Possibilistic c-Means model:

```{r}
model_sspcm <- fussclust::SSPCM(X = X, C = 3, alpha = 1, superF = superF)
```

Compare the estimated prototypes.

```{r}
model_ssfcm$V
```

```{r}
model_sspcm$V
```


# Details

## $\Gamma$ in possibilistic approach

The vector of cluster-specific hyperparameters $\Gamma = (\gamma_1, \ldots, \gamma_c)$
is specific to the possibilistic approach.

This is an optional argument in both `fussclust::PCM` and `fussclust:SSPCM`.
If user does not provide one, by default a unit vector $\Gamma = (1, \ldots, 1)$ is used.

However, the user may also choose to apply the initialization strategy described 
by [Krishnapuram and Keller (1993)](https://doi.org/10.1109/91.227387), which relies
on calculating:

$$\gamma_k \sim \frac{\sum_{j=1}^N u^2_{jk} d^2_{jk}}{\sum_{j=1}^N u^2_{jk}}$$

where the values in the formula are obtained by the initial run of Fuzzy c-Means.

This approach can be applied in `fussclust` by setting the argument `initFCM = TRUE`.

```{r, eval = FALSE}
model_pcm_init <- fussclust::PCM(X = X, C = 2, initFCM = TRUE)
```

## Retrieving intermediate results

The user may be interested not only in the final estimated memberships and prototypes
(that can be retrieved by `model$U` or `model$V`),
but also in the interim results.
Recall that the AO algorithm is iterative in its nature.

The history of three matrices can be investigated:

* Estimated memberships `\hat{U}^{(l)}`, accessed by `model$U_history`,
* Estimated prototypes `\hat{U}^{(l)}`, accessed by `model$V_history`,
* weights `\hat{\Phi}^{(l)}`, accessed by `model$Phi_history`.

By default, the values of these lists are set to `NULL`.

```{r}
model_fcm$U_history
```

To enable storing such results, the argument `story_history = TRUE` must be set accordingly.

```{r}
model_fcm_history <- fussclust::FCM(X = X, C = 2, store_history = TRUE)

cat(paste0(
  "Class of `model_fcm_history$U_history` is ", class(model_fcm_history$U_history),
  " and the length of the object is ", length(model_fcm_history$U_history),
  " since there were ", model_fcm_history$counter, " loops executed.",
  collapse = " "))
```













